Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-06 _paper.rmd
Working through Di's comments.

---
title: "Exploring Local Explanations of Non-linear Models Using Animated Linear Projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
#- name: "Nicholas Spyrison"
#   address: |
#     | ORCiD: 0000-0002-8417-0212
#   email: \email{nicholas.spyrison@monash.edu}
#   orcid: 0000-0002-8417-0212
#- name: "Dianne Cook"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 000-0002-3813-7155
#   orcid: 000-0002-3813-7155
#- name: "Kimbal Marriott"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 0000-0002-9813-0377
#   orcid: 0000-0002-9813-0377
abstract: |
 Artificial Intelligence (AI) has seen a revitalization in recent years with the prevalence of increasingly hard-to-interpret black-box models. The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear variable importance in the vicinity of each observation. These are simply linear projections, and as such can be further explored interactively to better understand the interaction between variables used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (footbal salaries, house prices) response variables. The methods are implemented in the __R__ package __cheem__, available on CRAN.
# documentclass: jdssv
#classoption: article
output:
  #bookdown::pdf_document2:
  pdf_document:
    keep_tex: true
    toc: false
    #template: template/jdssv_template.tex
    number_sections: yes
    fig_caption: yes
bibliography: cheem_paper.bib
#appendix:
# title:
#   formatted: "Methods for understanding the variable importance of local explanations of black-box models"
#   plain:     "Methods for understanding the variable importance of local explanations of black-box models"
#   short:     "cheem: radial tour scrutinizing local explanations"
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
 - \usepackage{setspace} ## for title page spacing
 - \usepackage{hyperref} ## for all sorts of linking
---

```{r loadlibraries, include=FALSE, cache=FALSE}
library("knitr")
library("kableExtra")
library("magrittr")
## Work packages
library("cheem")
library("spinifex")
## chunk options
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = FALSE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = TRUE,
  cache.lazy = FALSE)
```

<!-- # cheat sheet {#sec:cheatsheet} -->
<!-- A bib reference [@wickham_visualizing_2015]. -->
<!-- A [Section intro](#sec:cheatsheet) reference, alternatively, section \@ref(sec:intro) (with @; \\@ref(sec:intro)). -->
<!-- ```{r crest, echo=FALSE, out.height = "10%", out.width = "10%", fig.cap = "A caption for crest figure"} -->
<!-- knitr::include_graphics("./figures/crest.jpg") -->
<!-- ``` -->
<!-- A figure \@ref(fig:crest) reference (with @; \\@ref(fig:crest). -->
<!-- (ref:baseplotcap) Separate caption created above the R chunk -->
<!-- ```{r baseplot, echo=F, fig.cap = "(ref:baseplotcap)"} -->
<!-- plot(1,2) -->
<!-- ``` -->

@@

XXX DC Note that we need to make sure that the paper is readable for a machine learner, so a translation of terms is important. I've done this by parenthesizing the ML term in this introduction, which should be enough.

@@

<!-- Paper structure -->
The remainder of this paper is organized as follows. The following section, [Local explanations](#sec:explanations) covers the background of the local explanation, and the traditional visuals produced. [Tours and the radial tour](#sec:tour) explains the animations of continuous linear projections. The section [Application Design](#sec:applicationdesign) discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. The section [Case Studies](#sec:casestudies) illustrates the application to supervised learning with categorical and quantitative response variables. We conclude with a [Discussion](#sec:cheemdiscussion) of the insights gained and directions that might be explore in the future. (XXX DC Can you change the references to number of the section not title?)

@@

<!-- ## SHAP SHAP and history -->
SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation conceptually by examining the effect of other variables on the contribution of the variable of interest on predicting the response. This explanations almost all point to @shapley_value_1953's method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. @strumbelj_efficient_2010 introduced the use of SHAP for local explanations in ML models. However, it is also related to partial dependence plots (XXX Needs reference), used to explain the effect of a variable by predicting the response for a range of values on this variable, after fixing the value of all other variables to their mean. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015, which helps to understand the relative importance of each variable in the context of all other candidate variables.

@@

<!-- Fifa example -->
XXX DC does this example come from/motivated by https://ema.drwhy.ai/UseCaseFIFA.html ?? it needs to be cited here. 
To understand the use SHAP we use soccer data from FIFA 2020 season [@leone_fifa_2020]. There are 5000 observations of 9 skill measures (after aggregating highly correlated variables). A random forest model is fit with log wages, in 2020 Euros, as the response, using the skill measures as predictors. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \@ref(fig:shapdistrbd). We expect to see a difference in the attribution of the variable importance across the two positions of the players, which would be interpreted as how the player's salary depends on this combination of skill sets. Plot (b) is a modified breakdown plot [@gosiewska_ibreakdown_2019] where the order of variables is fixed, so the two observations can be more easily compared.

@@

XXX DC Can you remove "Variable" from the vertical axes? And remove reference to "permutations" in the figure text. Also, why include interaction and prediction - suggest removing these because there is no information. Perhaps switch the order, and put the breakdown plot first?

<!-- Illustration -->
<!-- Figure \@ref(fig:shapdistrbd) shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offense and movement skills are more important for the offensive player, while defensive and power skills are more informative to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel b) shows a simplified break-down plot [@gosiewska_ibreakdown_2019], where a local explanation is used to additively explain the difference from the intercept to the observation's prediction. Such additive approaches will show asymmetry in the variable ordering, so we opt to fix the order to panel a) by decreasing the sum of the SHAP values.-->

@@

<!-- ## Manual tours and its radial case-->

>>>>

---
title: "Exploring Local Explanations of Non-linear Models Using Animated Linear Projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
#- name: "Nicholas Spyrison"
#   address: |
#     | ORCiD: 0000-0002-8417-0212
#   email: \email{nicholas.spyrison@monash.edu}
#   orcid: 0000-0002-8417-0212
#- name: "Dianne Cook"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 000-0002-3813-7155
#   orcid: 000-0002-3813-7155
#- name: "Kimbal Marriott"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 0000-0002-9813-0377
#   orcid: 0000-0002-9813-0377
abstract: |
 Artificial Intelligence (AI) has seen a revitalization in recent years with the prevalence of increasingly hard-to-interpret black-box models. The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear variable importance in the vicinity of each observation. These are simply linear projections, and as such can be further explored interactively to better understand the interaction between variables used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (footbal salaries, house prices) response variables. The methods are implemented in the __R__ package __cheem__, available on CRAN.
# documentclass: jdssv
#classoption: article
output:
  ## bookdown::pdf_document2: allows \@ref()
  bookdown::pdf_document2:
  ## pdf_document: requires \ref{} 
  #pdf_document:
    keep_tex: true
    toc: false
    #template: template/jdssv_template.tex
    number_sections: yes
    fig_caption: yes
bibliography: cheem_paper.bib
#appendix:
# title:
#   formatted: "Methods for understanding the variable importance of local explanations of black-box models"
#   plain:     "Methods for understanding the variable importance of local explanations of black-box models"
#   short:     "cheem: radial tour scrutinizing local explanations"
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
 - \usepackage{setspace} ## for title page spacing
 - \usepackage{hyperref} ## for all sorts of linking
---

```{r loadlibraries, include=FALSE, cache=FALSE}
library("knitr")
library("kableExtra")
library("magrittr")
## Work packages
library("cheem")
library("spinifex")
## chunk options
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = FALSE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = TRUE,
  cache.lazy = FALSE)
```
<!-- bookdown::pdf_document2: allows \@ref(), pdf_document: requires \ref{} -->
<!-- # cheat sheet {#sec:cheatsheet} -->
<!-- A bib reference [@wickham_visualizing_2015]. -->
<!-- A [Section intro](#sec:cheatsheet) reference, alternatively, section \@ref(sec:intro) (with @; \\@ref(sec:intro)). -->
<!-- ```{r crest, echo=FALSE, out.height = "10%", out.width = "10%", fig.cap = "A caption for crest figure"} -->
<!-- knitr::include_graphics("./figures/crest.jpg") -->
<!-- ``` -->
<!-- A figure \@ref(fig:crest) reference (with @; \\@ref(fig:crest). -->
<!-- (ref:baseplotcap) Separate caption created above the R chunk -->
<!-- ```{r baseplot, echo=F, fig.cap = "(ref:baseplotcap)"} -->
<!-- plot(1,2) -->
<!-- ``` -->

@@

@@

<!-- Paper structure -->
The remainder of this paper is organized as follows. The following Section, \@ref(sec:explanations) covers the background of the local explanation, and the traditional visuals produced. Section \@ref(sec:tour) explains the animations of continuous linear projections. Section \@ref(sec:applicationdesign) discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. Then Section \@ref(#sec:casestudies) illustrates the application to supervised learning with categorical and quantitative response variables. We conclude with Section \@ref(sec:cheemdiscussion) of the insights gained and directions that might be explore in the future.

@@

<!-- ## SHAP and history -->
SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation conceptually by examining the effect of other variables on the contribution of the variable of interest on predicting the response. This explanations almost all point to @shapley_value_1953's method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. @strumbelj_efficient_2010 introduced the use of SHAP for local explanations in ML models. However, it is also related to partial dependence plots [@molnar_interpretable_2020], used to explain the effect of a variable by predicting the response for a range of values on this variable, after fixing the value of all other variables to their mean. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015, which helps to understand the relative importance of each variable in the context of all other candidate variables.

@@

<!-- Fifa example -->
Following the use case _Explanatory Model Analysis_ [@biecek_explanatory_2021], we use FIFA data to illustrate the use of SHAP. Consider soccer data from the FIFA 2020 season [@leone_fifa_2020]. There are 5000 observations of 9 skill measures (after aggregating highly correlated variables). A random forest model is fit regressing log wages [2020 Euros], from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \@ref(fig:shapdistrbd). We expect to see a difference in the attribution of the variable importance across the two positions of the players, which would be interpreted as how the player's salary depends on this combination of skill sets. Plot (b) is a modified breakdown plot [@gosiewska_ibreakdown_2019] where the order of variables is fixed, so the two observations can be more easily compared.

@@

@@

### 2022-01-06 _paper.tex
Working through Di's comments.

  pdfkeywords={R, XAI, explanable artificial intelligence, black-box
models, local explanations, grand tour, manual tour, radial tour},

@@

@@

Using a radial tour can be considered similar to counterfactual, what-if
analysis, such as \emph{ceteris paribus} (Biecek 2020). This phrase,
Latin for ``other things held constant'' or ``all else unchanged'',
shows how an observation's prediction would change from a marginal
change in one explanatory variable given that other variables are held
constant. It ignores correlations of the variables and imagines a case
that was not observed. In contrast, our approach is a geometric
explanation of the factual; it varies contributions of the variables by
rotating the basis, a reorientation of the data object. A constraint in
our approach is that the basis must remain orthonormal. That means, when
the contribution of one variable decreases, the contributions of others
necessarily increase such that there is a complete component in that
direction. This also ensures that what is seen is strictly a
low-dimensional projection from high-dimensions, and is thus an
interpretable visualisation.

The remainder of this paper is organized as follows. The following
section, \protect\hyperlink{sec:explanations}{Local explanations} covers
the background of the local explanation, and the traditional visuals
produced. \protect\hyperlink{sec:tour}{Tours and the radial tour}
explains the animations of continuous linear projections. The section
\protect\hyperlink{sec:applicationdesign}{Application Design} discusses
the visual layout in the interactive interface, how they facilitate
analysis, data preprocessing, and package infrastructure. The section
\protect\hyperlink{sec:casestudies}{Case Studies} illustrates the
application to supervised learning with categorical and quantitative
response variables. We conclude with a
\protect\hyperlink{sec:cheemdiscussion}{Discussion} of the insights
gained and directions that might be explore in the future. (XXX DC Can
you change the references to number of the section not title?)

@@

SHaply Additive exPlanations (SHAP) approximates the variable importance
in the vicinity of one observation conceptually by examining the effect
of other variables on the contribution of the variable of interest on
predicting the response. This explanations almost all point to Shapley
(1953)'s method to evaluate an individual's contribution to cooperative
games by permuting the players that contribute to the score. Strumbelj
and Kononenko (2010) introduced the use of SHAP for local explanations
in ML models. However, it is also related to partial dependence plots
(XXX Needs reference), used to explain the effect of a variable by
predicting the response for a range of values on this variable, after
fixing the value of all other variables to their mean. It could also be
considered to be similar to examining the coefficients from all subsets
regression, as described in Wickham, Cook, and Hofmann (2015), which
helps to understand the relative importance of each variable in the
context of all other candidate variables.

@@

XXX DC does this example come from/motivated by
\url{https://ema.drwhy.ai/UseCaseFIFA.html} ?? it needs to be cited
here. To understand the use SHAP we use soccer data from FIFA 2020
season (Leone 2020). There are 5000 observations of 9 skill measures
(after aggregating highly correlated variables). A random forest model
is fit with log wages, in 2020 Euros, as the response, using the skill
measures as predictors. We then extract the SHAP values of a star
offensive player (L. Messi) and defensive player (V. van Dijk). The
results are displayed in Figure @ref(fig:shapdistrbd). We expect to see
a difference in the attribution of the variable importance across the
two positions of the players, which would be interpreted as how the
player's salary depends on this combination of skill sets. Plot (b) is a
modified breakdown plot (Gosiewska and Biecek 2019) where the order of
variables is fixed, so the two observations can be more easily compared.

@@

XXX DC Can you remove ``Variable'' from the vertical axes? And remove
reference to ``permutations'' in the figure text. Also, why include
interaction and prediction - suggest removing these because there is no
information. Perhaps switch the order, and put the breakdown plot first?

@@

(ref:citeRf) Liaw and Wiener (2002) (ref:citeTs) Kominsarczyk et al.
(2021)

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to the color and shape of observations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [(ref:citeRf)] for compatibility with the local explanation, which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, compatible with tree-based models. Tree SHAP was calculated with the package \textbf{treeshap} [(ref:citeTs), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

@@

(ref:casepenguins) Species classification of Palmer penguin data. We
select a chinstrap penguin that is mislabeled as an adelie. By varying
the contribution to bill length, we observe the explanation does not
hold when bill length has a significant contribution. The .mp4 animation
of this tour can be found at
\href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_penguins.mp4}

@@

\caption{(ref:casepenguins)}\label{fig:casepenguins}
\end{figure}

In figure @ref(fig:casepenguins), a misclassified point is contrasted
with a correctly classified point of its observed class nearby in
data-space. The attribution space from the tree SHAP local explanations
is a more separable space, where the comparison is squarely in the
middle of the orange distribution. The primary observation is between
the predicted and observed clusters, a sign of uncertainty in the
prediction. The tour varies the contribution of bill length (b\_l) as
this variable differs most from the contribution of the comparison
observation. Downplaying the contribution of bill length is crucial to
the linear explanation of this observation being misclassified.

@@

(ref:casechocolates) Chocolates data type classification (milk or dark).
We select a chocolate labeled as dark though a random forest model
predicts it to be milk chocolate in light of the values on the
nutritional label. We vary the contribution of calories from fat.
Animated tour can be found at
\href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_chocolates.mp4}.

@@

\caption{(ref:casechocolates)}\label{fig:casechocolates}
\end{figure}

Figure @ref(fig:casechocolates) similarly shows that attribution-space
is more separable than data-space. Interestingly, the class imbalance
that we suspected was not observed; there are only six chocolates
labeled as dark and predicted as milk, while eight of the inverse case.
Calories from fat is the variable with the largest difference in
treeshap attribution between these points. In this case, it feels
strange to call the selected observation a misclassification of the
model. There are plausible reasons that a manufacturer has incentives to
cut corners and label their products different than what they are. This
feels more like a measurement theory-related problem. In this case, the
candy is being sold as dark chocolate, while nutritional value more
closely resembles milk chocolates.

@@

The 2020 season FIFA data (Leone 2020; Biecek 2018) contains many skill
measurements of soccer/football players and wage information. After
aggregation of the skill measurements, we regress the log wages {[}2020
euros{]} given just the skill aggregates. The model was fit from 5000
observations of the nine skill aggregates before being thinned to 500
players to mitigate occlusion and render time. We compare a leading
offensive fielder (L. Messi) with that of a top defensive fielder (V.
van Dijk), the same observations were used in figure
@ref(fig:shapdistrbd).

(ref:casefifa) FIFA 2020, regressing log wages {[}2020 Euros{]} from
aggregations of skill measurements. The primary observation is a star
offensive player (L. Messi) compared with a top defensive player (V. van
Dijk). The animate radial tour can be found at
\href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_fifa.mp4}

@@

\caption{(ref:casefifa)}\label{fig:casefifa}
\end{figure}

With figure @ref(fig:casefifa), we will test the premise of the local
explanation. If we remove reaction and movement skills from the basis,
offense skills are almost singularly important for explaining the
offensive player. We vary the contribution of offensive skills.
Offensive skills are removed in the tour (panel b, frame 3), and Messi
is no longer separated from the group. We also notice that accuracy has
rotated into the frame, maintaining some separability.

@@

(ref:caseames) Ames housing 2018 regressing log sales price {[}2018
USD{]}. Because the SHAP values are relatively well distributed across
the variables, there is a lot of redundant information to explain the
gap between select observations. This case may not be ideal for the
cheem methodology, though we did find a near singular frame in the
manual tour not, something we came looking for, but certainly, an
exciting feature to find. The corresponding animation is at
\href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_ames2018.mp4}

@@

\caption{(ref:caseames)}\label{fig:caseames}
\end{figure}

Figure @ref(fig:caseames) shows the global view and extrema of the tour.
The horizontal distance in the tour did not show a significant disparity
between our selected points. This is not particularly surprising as most
variables have a sizable contribution. Rotating any one variable out of
the frame will rotate other vital variables into the frame, preserving
most of the distance from intercept to prediction. However, the tour has
revealed an interesting feature worth discussing. Notice that the
observations pivot about the origin, the basis roughly halfway between
bases in frames one and two of panel b) the data is near a singular
profile. This means that there is a basis orthogonal to this point that
describes sizable variation. Knowing these singular bases can point
toward others with meaningful data variation.

@@

>>>>

  pdfkeywords={R, XAI, explanable artificial intelligence, black-box models, local explanations, grand tour, manual tour, radial tour},

@@

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

@@

Using a radial tour can be considered similar to counterfactual, what-if analysis, such as \emph{ceteris paribus} (Biecek 2020). This phrase, Latin for ``other things held constant'' or ``all else unchanged,'' shows how an observation's prediction would change from a marginal change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual; it varies contributions of the variables by rotating the basis, a reorientation of the data object. A constraint in our approach is that the basis must remain orthonormal. That means, when the contribution of one variable decreases, the contributions of others necessarily increase such that there is a complete component in that direction. This also ensures that what is seen is strictly a low-dimensional projection from high-dimensions, and is thus an interpretable visualisation.

The remainder of this paper is organized as follows. The following Section, \ref{sec:explanations} covers the background of the local explanation, and the traditional visuals produced. Section \ref{sec:tour} explains the animations of continuous linear projections. Section \ref{sec:applicationdesign} discusses the visual layout in the interactive interface, how they facilitate analysis, data preprocessing, and package infrastructure. Then Section @ref(\#sec:casestudies) illustrates the application to supervised learning with categorical and quantitative response variables. We conclude with Section \ref{sec:cheemdiscussion} of the insights gained and directions that might be explore in the future.

@@

SHaply Additive exPlanations (SHAP) approximates the variable importance in the vicinity of one observation conceptually by examining the effect of other variables on the contribution of the variable of interest on predicting the response. This explanations almost all point to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by permuting the players that contribute to the score. Strumbelj and Kononenko (2010) introduced the use of SHAP for local explanations in ML models. However, it is also related to partial dependence plots (Molnar 2020), used to explain the effect of a variable by predicting the response for a range of values on this variable, after fixing the value of all other variables to their mean. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each variable in the context of all other candidate variables.

@@

Following the use case \emph{Explanatory Model Analysis} (Biecek and Burzykowski 2021), we use FIFA data to illustrate the use of SHAP. Consider soccer data from the FIFA 2020 season (Leone 2020). There are 5000 observations of 9 skill measures (after aggregating highly correlated variables). A random forest model is fit regressing log wages {[}2020 Euros{]}, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). The results are displayed in Figure \ref{fig:shapdistrbd}. We expect to see a difference in the attribution of the variable importance across the two positions of the players, which would be interpreted as how the player's salary depends on this combination of skill sets. Plot (b) is a modified breakdown plot (Gosiewska and Biecek 2019) where the order of variables is fixed, so the two observations can be more easily compared.

@@

@@

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to the color and shape of observations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [Liaw and Wiener (2002)] for compatibility with the local explanation, which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, compatible with tree-based models. Tree SHAP was calculated with the package \textbf{treeshap} [Kominsarczyk et al. (2021), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

@@

@@

\caption{Species classification of Palmer penguin data. We select a chinstrap penguin that is mislabeled as an adelie. By varying the contribution to bill length, we observe the explanation does not hold when bill length has a significant contribution. The .mp4 animation of this tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_penguins.mp4}}\label{fig:casepenguins}
\end{figure}

In figure \ref{fig:casepenguins}, a misclassified point is contrasted with a correctly classified point of its observed class nearby in data-space. The attribution space from the tree SHAP local explanations is a more separable space, where the comparison is squarely in the middle of the orange distribution. The primary observation is between the predicted and observed clusters, a sign of uncertainty in the prediction. The tour varies the contribution of bill length (b\_l) as this variable differs most from the contribution of the comparison observation. Downplaying the contribution of bill length is crucial to the linear explanation of this observation being misclassified.

@@

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate in light of the values on the nutritional label. We vary the contribution of calories from fat. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_chocolates.mp4}.}\label{fig:casechocolates}
\end{figure}

Figure \ref{fig:casechocolates} similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. Calories from fat is the variable with the largest difference in treeshap attribution between these points. In this case, it feels strange to call the selected observation a misclassification of the model. There are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. This feels more like a measurement theory-related problem. In this case, the candy is being sold as dark chocolate, while nutritional value more closely resembles milk chocolates.

@@

The 2020 season FIFA data (Leone 2020; Biecek 2018) contains many skill measurements of soccer/football players and wage information. After aggregation of the skill measurements, we regress the log wages {[}2020 euros{]} given just the skill aggregates. The model was fit from 5000 observations of the nine skill aggregates before being thinned to 500 players to mitigate occlusion and render time. We compare a leading offensive fielder (L. Messi) with that of a top defensive fielder (V. van Dijk), the same observations were used in figure \ref{fig:shapdistrbd}.

@@

\caption{FIFA 2020, regressing log wages {[}2020 Euros{]} from aggregations of skill measurements. The primary observation is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). The animate radial tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_fifa.mp4}}\label{fig:casefifa}
\end{figure}

With figure \ref{fig:casefifa}, we will test the premise of the local explanation. If we remove reaction and movement skills from the basis, offense skills are almost singularly important for explaining the offensive player. We vary the contribution of offensive skills. Offensive skills are removed in the tour (panel b, frame 3), and Messi is no longer separated from the group. We also notice that accuracy has rotated into the frame, maintaining some separability.

@@

@@

\caption{Ames housing 2018 regressing log sales price {[}2018 USD{]}. Because the SHAP values are relatively well distributed across the variables, there is a lot of redundant information to explain the gap between select observations. This case may not be ideal for the cheem methodology, though we did find a near singular frame in the manual tour not, something we came looking for, but certainly, an exciting feature to find. The corresponding animation is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{github.com/nspyrison/cheem\_paper/blob/main/figures/case\_ames2018.mp4}}\label{fig:caseames}
\end{figure}

Figure \ref{fig:caseames} shows the global view and extrema of the tour. The horizontal distance in the tour did not show a significant disparity between our selected points. This is not particularly surprising as most variables have a sizable contribution. Rotating any one variable out of the frame will rotate other vital variables into the frame, preserving most of the distance from intercept to prediction. However, the tour has revealed an interesting feature worth discussing. Notice that the observations pivot about the origin, the basis roughly halfway between bases in frames one and two of panel b) the data is near a singular profile. This means that there is a basis orthogonal to this point that describes sizable variation. Knowing these singular bases can point toward others with meaningful data variation.

@@

\leavevmode\vadjust pre{\hypertarget{ref-molnar_interpretable_2020}{}}%
Molnar, Christoph. 2020. \emph{Interpretable Machine Learning}. Lulu. com.
